Newton-Type Methods for Non-Convex Optimization Under Inexact Hessian Information

### 2017-11-06 Paper-Presentation/main.tex
Complete PPT

\documentclass{beamer}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bookmark}
\usepackage{hyperref}
\newcommand{\h}{\nabla^{2}}
\newcommand{\g}{\nabla}
\newcommand{\xbold}{\mathbf{x}}
\newcommand{\sbold}{\mathbf{s}}
\newcommand{\mineig}{\lambda_{\min}}
\newcommand{\maxeig}{\lambda_{\max}}
%\documentclass{article}
%\usepackage{beamerarticle}
\usetheme{CambridgeUS}

@@

\begin{frame}{Definitions}
\begin{alertblock}{\((\epsilon_{g}, \epsilon_{h})\)-Optimality}
\label{sec-opt}
Given \(\epsilon_{g}, \epsilon_{h} \in (0, 1)\), \(\mathbf{\hat{x}}\) is an \((\epsilon_{g}, \epsilon_{h})\)-optimal solution to the problem \ref{main-obj}, if:
\begin{itemize}
\item \(||\g F(\mathbf{\hat{x}}) || \leq \epsilon_{g}\)
\item \(\mineig (\h F(\mathbf{\hat{x}})) \geq -\epsilon_{h}\)
\end{itemize}
\end{alertblock}
\pause
Note: If the function \(F\) satisfies \underline{strict saddle property}, then the above defined optimality condition ensures ``closeness'' to a local-minimum. % add citations
\pause
\begin{alertblock}{Lipschitz and Bounded Hessian (Hessian Regularity)}
Given \(\xbold_{t}\) and \(\sbold_{t}\) the \(t^{th}\) iterate and \(t^{th}\) update step respectively:
\begin{itemize}
\item \(||\h F(\xbold) - \h F(\xbold_{t})|| \leq L||\xbold - \xbold_{t}|| \hspace{4mm} \forall \xbold \in [\xbold_{t}, \xbold_{t} + \sbold_{t}]\)
\item \(||\h F(\xbold_{t}) || \leq B, \hspace{4mm} B < \infty\)
\end{itemize}
\end{alertblock}
\end{frame}

@@

\section{Finite-Sum Minimization}
\begin{frame}{Randomized Sub-sampling}
\begin{itemize}
\item<1->{Consider a modified unconstrained optimization problem
  \begin{equation}
    \min_{\xbold \in \mathbb{R}^{d}} F(\xbold) \triangleq \frac{1}{n}\displaystyle \sum_{i=1}^{n} f_{i}(\xbold)
  \end{equation}
where each \(f_{i}\) is smooth, but possibly non-convex.}
\item<2->{Seen in Machine Learning problems, where each \(f_{i}\) corresponds to the loss or mis-fit due to the \(i^{th}\) sample in the mini-batch/batch}
\item<3->{Large-scale problems have \(n, d >> 1\).}
\item<4->{Consider a probability distribution over \({1, 2, \ldots, n}\), denoted by \(\mathbf{p} = \{p_{i}\}_{i=1}^{n}\) and \(Pr(X = i) = p_{i} > 0, \hspace{2mm} \sum p_{i} = 1\). Based on \(\mathbf{p}\), we pick a sample \(\mathcal{S}\).}
\item<5->{Define the sub-sampled Hessian to be:
  \begin{equation}
    H(\xbold) \triangleq \frac{1}{n|\mathcal{S}|}\sum_{j \in \mathcal{S}} \frac{1}{p_{j}} \h f_{j}(\xbold)
  \end{equation}
}
\end{itemize}
\end{frame}

\begin{frame}{Randomized Sub-Sampling}
\begin{theorem}
Consider \(\displaystyle \sup_{\xbold\in\mathbb{R}^{d}} ||\h f_{i}(\xbold)|| \leq K_{i}, \hspace{2mm} \forall i \in \{1, 2, \ldots, n\}\). Let \(\displaystyle K_{\max} = \max_{i=1,\ldots, n} K_{i}\) and \(\hat{K} = \frac{1}{n} \displaystyle \sum_{i=1}^{n} K_{i}\). If \(p_{i} = \frac{1}{n} \hspace{1mm} \forall i \in \{1, 2, \ldots, n\}\), and \(|\mathcal{S}| \geq \frac{16K^{2}_{max}}{\epsilon^{2}}\log\frac{2d}{\delta} \), then with probability at least \(1 - \delta\), we have:
\begin{equation}
Pr\left(||H(\xbold) - \h F(\xbold)|| \leq \epsilon\right) \geq 1 - \delta
\end{equation}
\end{theorem}
\pause
\uncover<2->{Two major implications:}
\begin{itemize}
\item<3->{Higher the value of \(d\), more larger sample set is required. But this dependence is logarithmic.}
\item<4->{Higher the number of samples, lower the value of \(\epsilon\), in turn better ``inexact'' Hessian.}
\end{itemize}
\end{frame}
\end{document}

>>>>

\documentclass[10pt]{beamer}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bookmark}
\usepackage{hyperref}
\newcommand{\h}{\nabla^{2}}
\newcommand{\g}{\nabla}
\newcommand{\xbold}{\mathbf{x}}
\newcommand{\sbold}{\mathbf{s}}
\newcommand{\mineig}{\lambda_{\min}}
\newcommand{\maxeig}{\lambda_{\max}}
%\documentclass{article}
%\usepackage{beamerarticle}
\usetheme{CambridgeUS}

@@

\begin{frame}{Definitions}
\begin{alertblock}{\((\epsilon_{g}, \epsilon_{h})\)-Optimality}
\label{sec-opt}
Given \(\epsilon_{g}, \epsilon_{h} \in (0, 1)\), \(\mathbf{\hat{x}}\) is an \((\epsilon_{g}, \epsilon_{h})\)-optimal solution to the problem \ref{main-obj}, if:
\begin{itemize}
\item \(||\g F(\mathbf{\hat{x}}) || \leq \epsilon_{g}\)
\item \(\mineig (\h F(\mathbf{\hat{x}})) \geq -\epsilon_{h}\)
\end{itemize}
\end{alertblock}
\pause
Note: If the function \(F\) satisfies \underline{strict saddle property}, then the above defined optimality condition ensures ``closeness'' to a local-minimum. % add citations
\pause
\begin{alertblock}{Lipschitz and Bounded Hessian (Hessian Regularity)}
Given \(\xbold_{t}\) and \(\sbold_{t}\) the \(t^{th}\) iterate and \(t^{th}\) update step respectively:
\vspace{-2mm}
\begin{itemize}
\item \(||\h F(\xbold) - \h F(\xbold_{t})|| \leq L||\xbold - \xbold_{t}|| \hspace{4mm} \forall \xbold \in [\xbold_{t}, \xbold_{t} + \sbold_{t}]\)
\item \(||\h F(\xbold_{t}) || \leq B, \hspace{4mm} B < \infty\)
\end{itemize}
\end{alertblock}
\end{frame}

@@

\section{Finite-Sum Minimization}
\subsection{Random Sub-sampling}
\begin{frame}{General problem}
\begin{itemize}
\item<1->{Consider a modified unconstrained optimization problem
  \begin{equation}
    \min_{\xbold \in \mathbb{R}^{d}} F(\xbold) \triangleq \frac{1}{n}\displaystyle \sum_{i=1}^{n} f_{i}(\xbold)
  \end{equation}
where each \(f_{i}\) is smooth, but possibly non-convex.}
\item<2->{Seen in Machine Learning problems, where each \(f_{i}\) corresponds to the loss or mis-fit due to the \(i^{th}\) sample in the mini-batch/batch}
\item<3->{Large-scale problems have \(n, d >> 1\).}
\item<4->{Consider a probability distribution over \({1, 2, \ldots, n}\), denoted by \(\mathbf{p} = \{p_{i}\}_{i=1}^{n}\) and \(Pr(X = i) = p_{i} > 0, \hspace{2mm} \sum p_{i} = 1\). Based on \(\mathbf{p}\), we pick a sample \(\mathcal{S}\).}
\item<5->{Define the sub-sampled Hessian to be:
  \begin{equation}
    H(\xbold) \triangleq \frac{1}{n|\mathcal{S}|}\sum_{j \in \mathcal{S}} \frac{1}{p_{j}} \h f_{j}(\xbold)
  \end{equation}
}
\end{itemize}
\end{frame}

\begin{frame}{General problem - Uniform sampling of the Hessian}
\begin{theorem}
Consider \(\displaystyle \sup_{\xbold\in\mathbb{R}^{d}} ||\h f_{i}(\xbold)|| \leq K_{i}, \hspace{2mm} \forall i \in \{1, 2, \ldots, n\}\). Let \(\displaystyle K_{\max} = \max_{i=1,\ldots, n} K_{i}\). If \(p_{i} = \frac{1}{n} \hspace{1mm} \forall i \in \{1, 2, \ldots, n\}\), and \(|\mathcal{S}| \geq \frac{16K^{2}_{max}}{\epsilon^{2}}\log\frac{2d}{\delta} \), then with probability at least \(1 - \delta\), we have:
\begin{equation}
Pr\left(||H(\xbold) - \h F(\xbold)|| \leq \epsilon\right) \geq 1 - \delta
\end{equation}
\end{theorem}
\pause
\uncover<2->{Some major implications:}
\begin{itemize}
\item<3->{Higher the value of \(d\), more larger sample set is required. But this dependence is logarithmic.}
\item<4->{Higher the number of samples, lower the value of \(\epsilon\), in turn better ``inexact'' Hessian.}
\end{itemize}
\end{frame}

\begin{frame}{Special sub-problem}
\begin{itemize}
\item<1->{Applying a minor modification \(\xbold \rightarrow \mathbf{a}_{i}^{T}\xbold\), redefine the unconstrained optimization problem
  \begin{equation}
    \min_{\xbold \in \mathbb{R}^{d}} F(\xbold) \triangleq \frac{1}{n}\displaystyle \sum_{i=1}^{n} f_{i}(\mathbf{a}_{i}^{T}\xbold)
  \end{equation}
where each \(f_{i}\) is smooth, but possibly non-convex.}
\item<2->{Provides a more clear analogy to Machine Learning problems, where \(\mathbf{a}_{i}\) is a ``data-vector" e.g., Linear Kernel SVMs, Logistic Regression}
\item<3->{\(\h F(\xbold) = \frac{1}{n} \displaystyle \sum_{i=1}^{n} f_{i}''(\mathbf{a}_{i}^{T}\xbold) \mathbf{a}_{i}\mathbf{a}_{i}^{T}\)}
\end{itemize}
\end{frame}

\begin{frame}{Special sub-problem - Non-uniform sampling of the Hessian}
\begin{theorem}
Consider \(\displaystyle \sup_{\xbold\in\mathbb{R}^{d}} ||\h f_{i}(\xbold)|| \leq K_{i}, \hspace{2mm} \forall i \in \{1, 2, \ldots, n\}\). Let \(\hat{K} = \frac{1}{n}\sum_{i=1}^{n} K_{i}\). If \(p_{i} = \displaystyle \frac{|f_{i}''(\mathbf{a}_{i}^{T}\xbold)| ||\mathbf{a}_{i}||^{2}_{2}}{N}\), where \(N\) is a normalizing factor and \(|\mathcal{S}| \geq \frac{4\hat{K}^{2}}{\epsilon^{2}}\log\frac{2d}{\delta}\), then with probability at least \(1 - \delta\), we have:
\begin{equation}
Pr\left(||H(\xbold) - \h F(\xbold)|| \leq \epsilon\right) \geq 1 - \delta
\end{equation}
\end{theorem}
\pause
\uncover<2->{Some major implications:}
\begin{itemize}
\item<3->{Higher the value of \(d\), more larger sample set is required. But this dependence is logarithmic.}
\item<4->{Higher the number of samples, lower the value of \(\epsilon\), in turn better ``inexact'' Hessian.}
\item<5->{Probabilities are also determined by the ``length" of the data.}
\end{itemize}
\end{frame}

\begin{frame}{Usefulness and general ideas presented by these bounds}
\begin{itemize}
\item<1->{Both bounds depend on \(d\) in a logarithmic sense.}
\item<2->{Larger the size of \(\mathcal{S}\), better the ``inexact" Hessian.
  \begin{itemize}
    \item<3->{\(\hat{K} \leq K_{\max}\). So?}
    \item<4->{Smaller sample size \(|\mathcal{S}|\) required to achieve same level of accuracy in non-uniform sampling compared to uniform sampling}
    \item<5->{Proves helpful if distribution of \(K_{i}'s\) is skewed i.e., \(\hat{K} << K_{\max}\)}
  \end{itemize}
}
\item<3->{Choice of samples depend on the amount of curvature \(f_{i}''(\mathbf{a}_{i}^{T}\xbold)\), higher curvature \(\implies\) higher probability of being sampled.}
\end{itemize}
\end{frame}

\subsection{Convergence Analysis of Finite-Sum Minimization}
\begin{frame}{Probabilistic Convergence for Trust Region on Finite-Sum Minimization}
\begin{theorem}
Consider \(0 < \epsilon_{g}, \epsilon_{h} < 1\).
\begin{itemize}
\item<1->{\(H(\xbold)\) satisfies Inexact Hessian Regularity with \(\epsilon < (1- \eta)\nu\epsilon_{h}\)}
\item<2->{\(\h F(\xbold)\) satisfies Hessian regularity}
\item<3->{\(\delta_{0} = \delta\min\{\epsilon_{g}^{2}\epsilon_{h}, \epsilon_{h}^{3}\}\) for some \(\delta \in (0, 1)\)}
\item<4->{\(\mathcal{S}\) sampled with parameters \(\epsilon, \delta_{0}\) using either uniform or non-uniform sampling.}
\item<5->{if the approximate solution to \ref{TR} satisfies Sufficient Descent Condition-1, \uncover<6->{then the algorithm terminates in at most \(T = O\left(\max\{\epsilon_{h}^{-1}\epsilon_{g}^{-2}, \epsilon_{h}^{-3}\}\right)\) iterations, and with probability \(1 - \delta\), \(||\g F(\xbold_{T})|| \leq \epsilon_{g} \hspace{4mm} \text{ and } \hspace{4mm} \mineig(\h F(\xbold_{T})) \geq -(\epsilon + \epsilon_{h})\).}}
\end{itemize}
\end{theorem}
\end{frame}

\begin{frame}{Probabilistic Convergence for Cubic Regularization on Finite-Sum Minimization}
\begin{theorem}
Consider \(0 < \epsilon_{g}, \epsilon_{h} < 1\).
\begin{itemize}
\item<1->{\(H(\xbold)\) satisfies Inexact Hessian Regularity with \(\epsilon = \min\{\epsilon_{0}, \zeta\epsilon_{g}\}, \zeta \in (0, 0.5)\)}
\item<2->{\(\h F(\xbold)\) satisfies Hessian regularity}
\item<3->{\(\delta_{0} = \delta\min\{\epsilon_{g}^{3/2}, \epsilon_{h}^{3}\}\) for some \(\delta \in (0, 1)\)}
\item<4->{\(\mathcal{S}\) sampled with parameters \(\epsilon, \delta_{0}\) using either uniform or non-uniform sampling.}
\item<5->{if the approximate solution to \ref{CR} satisfies Sufficient Descent Condition-2 and Sufficient Descent for Optimal Complexity, \uncover<6->{then the algorithm terminates in at most \(T = O\left(\max\{\epsilon_{g}^{-3/2}, \epsilon_{h}^{-3}\}\right)\) iterations, and with probability \(1 - \delta\), \(||\g F(\xbold_{T})|| \leq \epsilon_{g} \hspace{4mm} \text{ and } \hspace{4mm} \mineig(\h F(\xbold_{T})) \geq -(\epsilon + \epsilon_{h})\).}}
\end{itemize}
\end{theorem}
\end{frame}

\section{Conclusion}
\begin{frame}{Summary}
\begin{itemize}
\item<1->{Approximation of curvature information}
\item<2->{Introduction to Trust Region and Cubic Regularization}
\item<3->{Achieving approximate optimality based on conditions or properties of the inexact Hessian}
\item<4->{Introduction to subsampling strategies}
\item<5->{Probabilistic convergence for Trust Region and Cubic Regularization on the finite-sum minimization problem}
\item<6->{Very theoretical - please view \href{arxiv.org/abs/1708.07827}{\beamergotobutton{Second-Order Optimization for Non-Convex Machine Learning: An Empirical Study}}}
\end{itemize}
\end{frame}

\begin{frame}{Questions??}
\end{frame}
\end{document}
